Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-08-20 Manuscript/manuscriptPS.tex

Calcium binding affinity to membranes containing negatively charged PS lipids can be
experimentally quantified by measuring the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the measurement is not compromised by the dehydrated lipid--ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating too strong calcium binding affinity.
The sole exception is the CHARMM36 model when paired with the NBfix
correction for calcium~\cite{kim16}. For this combination, the modulation of order parameters is underestimated, 
indicating a weaker binding affinity than in experiments.
Notably, the CHARMM36 model together with the NBfix corrections \cite{venable13,kim16} suggest similar binding affinities of
calcium and sodium for a POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
is underestimated in the CHARMM36 model when using the NBfix for calcium \cite{kim16}, but overestimated 
in all the other tested models. This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

>>>>

Calcium binding affinity to membranes containing negatively charged PS lipids can be
experimentally quantified by measuring the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the measurement is not compromised by the dehydrated lipid--ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating too strong calcium binding affinity.
The sole exception is the CHARMM36 model when paired with the NBfix
corrections for calcium~\cite{kim16,han2018graph}. For these combinations, the modulation of order parameters is underestimated, 
indicating a weaker binding affinity than in experiments.
Notably, the CHARMM36 model together with the NBfix corrections \cite{venable13,kim16} suggest similar binding affinities of
calcium and sodium for a POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
is underestimated in the CHARMM36 model when using the NBfix for calcium \cite{kim16,han2018graph}, but overestimated 
in all the other tested models. This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).
